Identifying and correcting for misspecifications\\in GWAS summary statistics and polygenic scores

### 2022-07-20 paper/paper-misspec.tex
last minor

We form nearly independent LD blocks using the optimal algorithm developed in \citet{prive2021optimal}.
{\color{red}Note that a new parameter \texttt{max\_r2} has been added to this method since, which controls the maximum single squared correlation allowed outside blocks, which offers an extra guarantee that the splitting is good and makes the function much faster by discarding lots of possible splits.
For different numbers of blocks and maximum number of variants in each block, we use the split that minimizes $\big(C2 \cdot \sqrt{5 + C1}\big)$, where $C1$ is the sum of squared correlations outside the blocks, and $C2$ is the sum of squared sizes of the blocks (Figure \ref{fig:cost-split}).
Twelve blocks from chromosome 22 are used in the simulations, and between 731 and 2527 in the real data analyses (cf.\ next section).}
Having a correlation matrix with independent blocks prevents small errors in the algorithm (e.g.\ the Gibbs sampler in LDpred2) from propagating to too many variants.
It also makes running LDpred2 (and lassosum2) faster, taking e.g.\ 70\% of the initial time (since only 70\% of the initial non-zero values of the correlation matrix are kept).
{\color{red}Note that we also compute the correlations within the same blocks and from the same data for PRS-CS.}

@@

\citet{marchini2010genotype} showed that the IMPUTE INFO measure is highly concordant with the MACH measure $\frac{\text{var}(G_j)}{2 \hat{\theta}_j (1 - \hat{\theta}_j)}$, where $\hat{\theta}_j$ is the estimated allele frequency of $G_j$, the genotypes for variant $j$.
Therefore, when using the expected genotypes from imputation (allele dosages), their standard deviations are often lower than the expected value under Hardy--Weinberg equilibrium, because $\text{INFO}_j \approx \frac{\text{var}(G_j)}{2 \hat{\theta}_j (1 - \hat{\theta}_j)}$.
In simulations (cf.\ Methods section ``Data for simulations''), we verify that $\text{sd}(G_j)^{\text{true}} \approx \text{sd}(G_j)^{\text{imp}} / \sqrt{\text{INFO}_j}$ (Figure \ref{fig:compare-sd}).
As a direct consequence of the lower standard deviation of dosages, we also show that GWAS effect sizes $\hat{\gamma}$ computed from imputed dosages are overestimated: $\hat{\gamma}_j^{\text{true}} \approx \hat{\gamma}_j^{\text{imp}} \cdot \sqrt{\text{INFO}_j}$ and $\text{se}(\hat{\gamma}_j)^{\text{true}} \approx \text{se}(\hat{\gamma}_j)^{\text{imp}} \cdot \sqrt{\text{INFO}_j}$ (Figures \ref{fig:compare-beta} and \ref{fig:compare-beta-se}).
This is the first correction of summary statistics we consider in the simulations below.
As a second option, instead of using dosages to compute the GWAS summary statistics, it has been argued that using multiple imputation (MI) would be more appropriate \cite[]{Palmer2016}.
{\color{red}Here MI consists in forming multiple (e.g.\ 20) datasets of hard-called genotypes by sampling them according to the imputation probabilities stored in the BGEN files, performing a GWAS on each of them, and pooling the GWAS estimates.}
In simulations, we show that $\hat{\gamma}_j^{\text{MI}} \approx \hat{\gamma}_j^{\text{imp}} \cdot \text{INFO}_j$ and $Z_j^{\text{MI}} \approx Z_j^{\text{imp}} \cdot \text{INFO}_j$, where $Z = \hat{\gamma} / \text{se}(\hat{\gamma})$ (Figure \ref{fig:compare-mi}).
This is the second correction of summary statistics we implement in the simulations below, along with $n_j \cdot \text{INFO}_j$ as the new per-variant sample sizes. 
Finally, we consider an in-between solution as a third correction, using $\hat{\gamma}_j = \hat{\gamma}_j^{\text{imp}} \cdot \text{INFO}_j$, $\text{se}(\hat{\gamma}_j) = \text{se}(\hat{\gamma}_j)^{\text{imp}} \cdot \sqrt{\text{INFO}_j}$, and $n_j \cdot \text{INFO}_j$ as the per-variant sample sizes. 
Note that we have recomputed INFO scores for the subset of 362,307 European individuals used in this paper since they can differ substantially from the ones reported by the UK Biobank for the whole data (Figures \ref{fig:compare-info} and \ref{fig:compare-maf-info}).

@@

Based on previous results from simulations and real-data analyses, we use this section to suggest some recommendations on how to handle misspecifications in GWAS summary statistics and to get the most of polygenic scores. We recall that an overview of these misspecifications, along with possible consequences, and possible corrections is summarized in table \ref{tab:overview}. First, standard quality control should always be performed, e.g.\ variants with low frequency or low imputation quality should be removed; just make sure to recompute MAF and INFO scores for the homogeneous subset of individuals you are interested in. Second, comparison of allele frequencies can be used to e.g.\ detect problems with signs of effects, while comparing standard deviations can be used to detect other issues such as low per-variant GWAS sample sizes.
Third, the LD reference panel should be as close as possible (in terms of genetic ancestry) as the data used to derive the GWAS. In case of any doubt, ancestry proportions can be estimated from GWAS allele frequencies using the method derived in \citet{prive2021using}.
Fourth, if some validation data is available for tuning hyper-parameters, one can use methods such as lassosum2 and LDpred2(-grid). The grid of parameters used in lassosum2 includes a broad range of regularization, which makes it very robust. For LDpred2, we recommend to also include a smaller value for the heritability in the grid of parameters, to allow for more regularization when needed. As these two methods use the same data and the same principle (i.e.\ hyper-parameter tuning), one could run both methods and should the best one when tuning hyper-parameters.
Fifth, if no validation/tuning dataset is available, we recommend using LDpred2-auto with the two new parameters introduced here, or PRS-CS-auto if there are strong misspecifications and the phenotype of interest is known to have small effects only. 
Sixth, we recommend to form independent LD blocks in the 
LD matrix as it proved to make the methods more robust (and slightly faster).
Most of these new recommendations are included in the LDpred2 tutorial.}

>>>>

We form nearly independent LD blocks using the optimal algorithm developed in \citet{prive2021optimal}.
{\color{red}Note that a new parameter \texttt{max\_r2} has now been added to this method, which controls the maximum single squared correlation allowed outside blocks, which offers an extra guarantee that the splitting is good and makes the function much faster by discarding lots of possible splits.
For different numbers of blocks and maximum number of variants in each block, we use the split that minimizes $\big(C2 \cdot \sqrt{5 + C1}\big)$, where $C1$ is the sum of squared correlations outside the blocks, and $C2$ is the sum of squared sizes of the blocks (Figure \ref{fig:cost-split}).
Twelve blocks from chromosome 22 are used in the simulations, and between 731 and 2527 in the real data analyses (cf.\ next section).}
Having a correlation matrix with independent blocks prevents small errors in the algorithm (e.g.\ the Gibbs sampler in LDpred2) from propagating to too many variants.
It also makes running LDpred2 (and lassosum2) faster, taking e.g.\ 70\% of the initial time (since only 70\% of the initial non-zero values of the correlation matrix are kept).
{\color{red}Note that we also compute the correlations within the same blocks and from the same data for PRS-CS.}

@@

\citet{marchini2010genotype} showed that the IMPUTE INFO measure is highly concordant with the MACH measure $\frac{\text{var}(G_j)}{2 \hat{\theta}_j (1 - \hat{\theta}_j)}$, where $\hat{\theta}_j$ is the estimated allele frequency of $G_j$, the genotypes for variant $j$.
Therefore, when using the expected genotypes from imputation (allele dosages), their standard deviations are often lower than the expected value under Hardy--Weinberg equilibrium, because $\text{INFO}_j \approx \frac{\text{var}(G_j)}{2 \hat{\theta}_j (1 - \hat{\theta}_j)}$.
In simulations (cf.\ Methods section ``Data for simulations''), we verify that $\text{sd}(G_j)^{\text{true}} \approx \text{sd}(G_j)^{\text{imp}} / \sqrt{\text{INFO}_j}$ (Figure \ref{fig:compare-sd}).
As a direct consequence of the lower standard deviation of dosages, we also show that GWAS effect sizes $\hat{\gamma}$ computed from imputed dosages are overestimated: $\hat{\gamma}_j^{\text{true}} \approx \hat{\gamma}_j^{\text{imp}} \cdot \sqrt{\text{INFO}_j}$ and $\text{se}(\hat{\gamma}_j)^{\text{true}} \approx \text{se}(\hat{\gamma}_j)^{\text{imp}} \cdot \sqrt{\text{INFO}_j}$ (Figures \ref{fig:compare-beta} and \ref{fig:compare-beta-se}).
This is the first correction of summary statistics we consider in the simulations below.
As a second option, instead of using dosages to compute the GWAS summary statistics, it has been argued that using multiple imputation (MI) would be more appropriate \cite[]{Palmer2016}.
{\color{red}Here MI consists in forming multiple (e.g.\ 20) datasets of hard-called genotypes by sampling them according to the imputation probabilities stored in the BGEN files, performing a GWAS on each of these datasets, and pooling the GWAS estimates.}
In simulations, we show that $\hat{\gamma}_j^{\text{MI}} \approx \hat{\gamma}_j^{\text{imp}} \cdot \text{INFO}_j$ and $Z_j^{\text{MI}} \approx Z_j^{\text{imp}} \cdot \text{INFO}_j$, where $Z = \hat{\gamma} / \text{se}(\hat{\gamma})$ (Figure \ref{fig:compare-mi}).
This is the second correction of summary statistics we implement in the simulations below, along with $n_j \cdot \text{INFO}_j$ as the new per-variant sample sizes. 
Finally, we consider an in-between solution as a third correction, using $\hat{\gamma}_j = \hat{\gamma}_j^{\text{imp}} \cdot \text{INFO}_j$, $\text{se}(\hat{\gamma}_j) = \text{se}(\hat{\gamma}_j)^{\text{imp}} \cdot \sqrt{\text{INFO}_j}$, and $n_j \cdot \text{INFO}_j$ as the per-variant sample sizes. 
Note that we have recomputed INFO scores for the subset of 362,307 European individuals used in this paper since they can differ substantially from the ones reported by the UK Biobank for the whole data (Figures \ref{fig:compare-info} and \ref{fig:compare-maf-info}).

@@

Based on previous results from simulations and real-data analyses, we use this section to suggest some recommendations on how to handle misspecifications in GWAS summary statistics and to get the most of polygenic scores. We recall that an overview of these misspecifications, along with possible consequences, and possible corrections is summarized in table \ref{tab:overview}. First, standard quality control should always be performed, e.g.\ variants with low frequency or low imputation quality should be removed; just make sure to recompute MAF and INFO scores for the homogeneous subset of individuals you are interested in. Second, comparison of allele frequencies can be used to e.g.\ detect problems with signs of effects, while comparing standard deviations can be used to detect other issues such as low per-variant GWAS sample sizes.
Third, the LD reference panel should be as close as possible (in terms of genetic ancestry) as the data used to derive the GWAS. In case of any doubt, ancestry proportions can be estimated from GWAS allele frequencies using the method derived in \citet{prive2021using}.
Fourth, if some validation data is available for tuning hyper-parameters, one can use methods such as lassosum2 and LDpred2(-grid). The grid of parameters used in lassosum2 includes a broad range of regularization, which makes it very robust. For LDpred2, we recommend to also include a smaller value for the heritability in the grid of parameters, to allow for more regularization when needed. As these two methods use the same data and the same principle (i.e.\ hyper-parameter tuning), one could run both methods and choose the best one when tuning hyper-parameters.
Fifth, if no validation/tuning dataset is available, we recommend using LDpred2-auto with the two new parameters introduced here, or PRS-CS-auto if there are strong misspecifications and the phenotype of interest is known to have small effects only. 
Sixth, we recommend to form independent LD blocks in the 
LD matrix as it proved to make the methods more robust (and slightly faster).
Most of these new recommendations are included in the LDpred2 tutorial.}
